A Monte Carlo Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
Converted eps files to pdf files for compatibility with linux pdflatex.

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
\usepackage{subfigure}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

@@

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ mesh). Both the adjoint and direct solvers are used
    with the five point and nine point stencils. A significant CPU
    time speedup is noted with the adjoint method due to the reduced
    number of histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ mesh. Both the adjoint and direct solvers are used with
    the five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

@@

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.eps}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. Both
    methods give similar iteration behavior.}
  \label{fig:poisson_convergence}
\end{figure}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution steps:

@@

As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
describe each of these steps in turn.

@@

However, a more natural interpretation for Monte Carlo transport
practitioners is to use the machinery described in
\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
Functions (PDF) that describe the transport of particles between
cells.  We then do a Monte Carlo transport calculation using
Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
collision. Equation~(\ref{eq:adjoint-probability}) gives the
probability of scattering to an adjacent cell. We tally the
contribution of $\phi$ in each cell using
Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
relative weight cutoff that is defined in
Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these steps in
more detail.

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{marshak_10.eps}}
  \caption{
    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
    solution is compared to the MCSA solution.  Note that the MCSA,
    PCG, and PFIXED methods all give numerically identical answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{spectral_radius.eps}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_CPU.eps}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_Ns.eps}}
  \caption{
    Maximum number of stages per cycle versus number of particles per
    stage for the Marshak problem.  The number of particles is the
    requested number.  In each stage the exact number can vary
    depending upon the number of cells and source strength per cell.}
  \label{fig:CPU_Ns}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_wc_CPU.eps}}
  \caption{
    CPU time versus relative weight cutoff for the Marshak problem.
    Results are shown for 10 and 50 requested particles per stage.
    The numbers in parenthesis show the maximum number of particles
    run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mesh.eps}}
  \caption{Multi-material problem mesh.  The duct region is blue.  The
    wall is red.  The foil is green.  The mesh is $60\times 60\times
    60$.}
  \label{fig:multi-mat-mesh}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{T_multi_mat.eps}}
  \caption{
    Two-dimensional plot of the temperature at 100~sh on a cut-plane
    positioned at the mid-point of the $z$-axis.  The positions of the
    edit points where the time-evolution of the temperature is plotted
    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[htpb!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.eps}} &
      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.eps}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{subfigure}
%% or use the.pdffig package if you prefer to use the old commands
%% \usepackage.pdffig}

@@

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < .pdfilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ mesh). Both the adjoint and direct solvers are used
    with the five point and nine point stencils. A significant CPU
    time speedup is noted with the adjoint method due to the reduced
    number of histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ mesh. Both the adjoint and direct solvers are used with
    the five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

@@

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. Both
    methods give similar iteration behavior.}
  \label{fig:poisson_convergence}
\end{figure}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution s.pdf:

@@

As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
describe each of these s.pdf in turn.

@@

However, a more natural interpretation for Monte Carlo transport
practitioners is to use the machinery described in
\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
Functions (PDF) that describe the transport of particles between
cells.  We then do a Monte Carlo transport calculation using
Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
collision. Equation~(\ref{eq:adjoint-probability}) gives the
probability of scattering to an adjacent cell. We tally the
contribution of $\phi$ in each cell using
Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
relative weight cutoff that is defined in
Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these s.pdf in
more detail.

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{
    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
    solution is compared to the MCSA solution.  Note that the MCSA,
    PCG, and PFIXED methods all give numerically identical answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{spectral_radius.pdf}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
  \caption{
    Maximum number of stages per cycle versus number of particles per
    stage for the Marshak problem.  The number of particles is the
    requested number.  In each stage the exact number can vary
    depending upon the number of cells and source strength per cell.}
  \label{fig:CPU_Ns}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
  \caption{
    CPU time versus relative weight cutoff for the Marshak problem.
    Results are shown for 10 and 50 requested particles per stage.
    The numbers in parenthesis show the maximum number of particles
    run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mesh.pdf}}
  \caption{Multi-material problem mesh.  The duct region is blue.  The
    wall is red.  The foil is green.  The mesh is $60\times 60\times
    60$.}
  \label{fig:multi-mat-mesh}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{
    Two-dimensional plot of the temperature at 100~sh on a cut-plane
    positioned at the mid-point of the $z$-axis.  The positions of the
    edit points where the time-evolution of the temperature is plotted
    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[htpb!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.pdf}} &
      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.pdf}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}
